Keep your Laziness in Check

### 2018-03-07 paper/main.tex
[paper] finished some TODOs in section 1 and 2

How, then, can we test and catch such laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following example using the standard
\lk{take} function on lists, which accepts an integer $n$ and a list, and returns
a prefix of the list with length $n$. If $n$ is larger than the length
of the original list, then the entire list is returned.
%
\begin{inlinecode}
take_spec :: Shaped a => Spec '[Int, [a]] [a]
take_spec =
  Spec $ \predict d n xs ->
    predict n (if n > length xs then d else d ++ thunk)
\end{inlinecode}
%
\hzh{TODO: describe what this spec means in English}

@@

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the used part of a data
structure. For lists, \textit{demands} will be prefixes of lists. We
use the type \lk{ListD} to represent such demands.
%
\begin{inlinecode}
  data ListD a = NilD
               | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and \lk{T}, a nullary constructor,
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor. For example, given
the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)} represents
the \textit{demand} that reveals the first 2 \lk{Cons} constructors of
the list, but does not evaluate any other parts of the list. To
improve readability of demand values, we will also use the
syntax \lk{\_ : \_ : \_} where appropriate, the underscore $\lk{\_}$
represents a thunk.

In Haskell, to trigger the evaluation of a lazy function, we need to
exert some demand on the result of that function. This suggests a
natural flow for the specification of a lazy function --- given the
demand on the output, the specification should calculate the demands
on the input. However, this is not enough in general because the
demand on the input could depend on the \textit{value} of the inputs
as well. The \lk{take} function introduced in \Cref{sec:intro} is such
a case since the length of the output from \lk{take} depends on its
integer argument.

Putting these pieces together, we arrive at an appropriate type for
the specification of \lk{rotate}: \lk{ListD a -> [a] -> [a] -> [a] ->
(ListD a, ListD a, ListD a)}.
%
\hzh{The description here needs improvement}
To specify the laziness behavior of \lk{rotate}, we return to Okasaki's insight:
as long as the front list is longer than the back list, demanding the first
element from the result of \lk{rotate} only causes evaluation of the first
element of the front list ~\cite{Okasaki95simpleand}. Using this precondition
for \lk{rotate}, we arrive at the following executable
specification \lk{rotate\_spec}.

Turning our intuition above into a Haskell function, we complete the
specification of \lk{rotate} in \Cref{fig:rotate-spec}.
%
\begin{figure}[H]
\centering
\hzh{TODO}
\begin{inlinecode}
TODO: rewrite this spec so it looks like the invariant in Okasaki's paper.
\end{inlinecode}
\caption{Specification for \lk{rotate}}
\label{fig:rotate-spec}
\end{figure}
%
The \lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example. \hzh{should we show what a
counter example looks like?}\leo{Definitely!}

>>>>

How, then, can we test and catch such laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following example using the standard
\lk{take} function on lists, which accepts an integer $n$ and a list, and returns
a prefix of the list with length $n$. If $n$ is larger than the length
of the original list, then the entire list is returned.
%
\begin{inlinecode}
take_spec :: Shaped a => Spec '[Int, [a]] [a]
take_spec =
  Spec $ \predict d n xs ->
    predict n (if n > length xs then d else d ++ thunk)
\end{inlinecode}
%
The \lk{Spec} constructor takes a function, whose first
argument \lk{predict} is the predicted demand on the \lk{Int} argument
of \lk{take}, and second argument the predicted demand on the \lk{[a]}
argument. This spec simply says that the \lk{Int} argument will always
be used, and the demand on the list argument is either the same as the
demand on the output, or demand on the output plus some
unevaluated \lk{thunk} at the end. The former case happens when $n$ is
larger than the list's length, otherwise the latter case applies.

@@

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the used part of a data
structure. For lists, \textit{demands} will be prefixes of lists. We
use the type \lk{ListD} to represent such demands.
%
\begin{inlinecode}
  data ListD a = NilD
               | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and \lk{T}, a nullary constructor,
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor. For example, given
the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)} represents
the \textit{demand} that reveals the first 2 \lk{Cons} constructors of
the list, but does not evaluate any other parts of the list. To
improve readability of demand values, we will also use the
syntax \lk{\_ : \_ : \_} instead of explicit demand type values. In
this expression, the first two underscores represent unevaluated
elements in the list, while the last underscore represents unevaluated
tail of the list.

In Haskell, to trigger the evaluation of a lazy function, we need to
exert some demand on the result of that function. This suggests a
natural flow for the specification of a lazy function --- given the
demand on the output, the specification should calculate the demands
on the input. However, this is not enough in general because the
demand on the input could depend on the \textit{value} of the inputs
as well. The \lk{take} function introduced in \Cref{sec:intro} is such
a case.

Putting these pieces together, we can specify the laziness behavior
of \lk{rotate} with Okasaki's insight: as long as the front list is
longer than the back list, demanding the first element from the result
of \lk{rotate} only causes evaluation of the first element of the
front list ~\cite{Okasaki95simpleand}. Since \lk{rotate} is always
used by supplying the empty list \lk{[]} as the accumulator argument,
we define \lk{rot fs bs = rotate fs bs []}, and show a specification
of \lk{rot}.
%
\begin{inlinecode}
rot_spec :: Shaped a => Spec '[[a], [a]] [a]
rot_spec =
  Spec $ \predict d fs bs ->
    let (fs', bs') = splitAt (length fs) (cap d)
        spineLen  = length (cap (d ++ [undefined]))  -- # of spine thunks forced
        overflow  = spineLen       > length fs  -- begun taking from bs?
        overrot   = length (cap d) > length bs  -- forced all of bs?
        padLength =
          length bs `min`
            if overflow
            then length bs - length bs'
            else length (cap d)
        spinePad = replicate padLength thunk
    in predict
         (                    fs' ++ if overflow            then [] else thunk)
         (spinePad ++ reverse bs' ++ if overflow || overrot then [] else thunk)
\end{inlinecode}
%
The \lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example.
\begin{figure}[H]
\centering
\includegraphics[width=0.8\textwidth]{revappend_counterexample}
\caption{Counterexample for \lk{revAppend} under \lk{rot\_spec}}
\end{figure}
